CTCF threshold analysis

### 2022-11-14 07_CTCF_Threshold.Rmd
Overlap with cell specific CTCF sites analysis

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
# Color palette for the heatmap, https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
col3 <- colorRampPalette(c('blue', 'white', 'red'))(20)
# col3 <- colorRampPalette(c('blue', 'gray', 'yellow'))(20)
# col3 <- colorRampPalette(c('green', 'black', 'red'))(20)
# col3 <- colorRamps::green2red(n = 20)
library(GenomicRanges)
```

@@

# Input files
fileNameIn1 <- file.path(dir_project, "RData", "mm10.MA0139.1.bed")
fileNameIn2 <- file.path(dir_project, "data", "mm10-CTCF.bed")
# Output files
fileNameOut1 <- file.path(dir_project, "results", "Figure_mouse_pvalues_threshold.svg")
```

@@

ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
# dev.off()
# svg(filename = fileNameOut1, width = 7, height = 3)
# print(p)
# dev.off()

```

>>>>

```{r libraries}
library(tidyverse)
library(readxl)
library(writexl)
library(cowplot)
library(stringr)
library("ggsci")
library(scales)
# scales::show_col(pal_lancet("lanonc")(8))
mycols = pal_lancet("lanonc")(8)
# Color palette for the heatmap, https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
col3 <- colorRampPalette(c('blue', 'white', 'red'))(20)
# col3 <- colorRampPalette(c('blue', 'gray', 'yellow'))(20)
# col3 <- colorRampPalette(c('green', 'black', 'red'))(20)
# col3 <- colorRamps::green2red(n = 20)
library(GenomicRanges)
library(rtracklayer)
library(plyranges)
```

@@

# Input files
fileNameIn1 <- file.path(dir_project, "RData", "BEDfiles", "hg38.MA0139.1.bed")
fileNameIn2 <- file.path(dir_project, "data", "GRCh38-CTCF.bed")
fileNameIn3 <- file.path(dir_project, "data", "UCSC_CTCF.tsv")
# Output files
fileNameOut1 <- file.path(dir_project, "results", "Figure_human_pvalues_threshold.svg")
fileNameOut2 <- file.path(dir_project, "results", "overlap_meme_below.csv")
fileNameOut3 <- file.path(dir_project, "results", "overlap_meme_above.csv")
```

@@

ggsave(filename = fileNameOut1, width = 4.5, height = 1.5)
# dev.off()
# svg(filename = fileNameOut1, width = 7, height = 3)
# print(p)
# dev.off()
```

## Overlap with cell-specific sites

```{r}
# Separate FIMO CTCF peaks into above and below the threshold
pval_threshold <- 1E-6
gr_meme_above <- gr_meme %>% filter(pval <= pval_threshold) %>% keepStandardChromosomes(., pruning.mode = "tidy") %>% sort()
gr_meme_below <- gr_meme %>% filter(pval >  pval_threshold) %>% keepStandardChromosomes(., pruning.mode = "tidy") %>% sort()
print(paste("Number of CTCFs above", pval_threshold, "p-value threshold", length(gr_meme_above)))
print(paste("Number of CTCFs below", pval_threshold, "p-value threshold", length(gr_meme_below)))
# Get all cell type-specific CTCF sites
# IDs of all CTCF experiments
mtx <- read_tsv(fileNameIn3, col_names = FALSE)
# Session for UCSC broser
mySession <- browserSession()
genome_id <- "hg38"
genome(mySession) <- genome_id
# Overlap counts
overlap_meme_above <- cbind()
overlap_meme_below <- cbind()
for (i in 158:nrow(mtx)) {
  print(paste(i, "out of", nrow(mtx)))
  table_name <- paste0("encTfChipPk", mtx[i, "X4"])
  query <- ucscTableQuery(mySession, table = table_name)
  gr_ctcf <- getTable(query) %>% makeGRangesFromDataFrame(.) %>% keepStandardChromosomes(., pruning.mode = "tidy") %>% sort()
  overlap_meme_below <- cbind(overlap_meme_below, countOverlaps(gr_meme_below, gr_ctcf))
  overlap_meme_above <- cbind(overlap_meme_above, countOverlaps(gr_meme_above, gr_ctcf))
  gc()
  write_csv(as.data.frame(overlap_meme_below), fileNameOut2)
  write_csv(as.data.frame(overlap_meme_above), fileNameOut3)
}

overlap_meme_below <- read_csv(paste0(fileNameOut2, ".gz"))
dim(overlap_meme_below)
overlap_meme_above <- read_csv(paste0(fileNameOut3, ".gz"))
dim(overlap_meme_above)

# The proportion of CTCF sites detected at p-value > 1E-6 overlapping at least one experimentally detected cell type-specific CTCF sites was XX +/-YY.
# In contrast, this number for CTCF sites detected at p-value <= 1E-6 was XX+/-YY, suggesting higher proportion of false positives among less significant CTCF sites.
overlap_meme_above_summary <- rowSums(overlap_meme_above)
overlap_meme_below_summary <- rowSums(overlap_meme_below)
print(paste("Average number of times less significant CTCF sites overlap with experimental CTCF sites", mean(overlap_meme_below_summary), "+/-", sd(overlap_meme_below_summary)))
# [1] "Average number of times less significant CTCF sites overlap with experimental CTCF sites 7.43024371211658 +/- 29.2354917718536"
print(paste("Average number of times more significant CTCF sites overlap with experimental CTCF sites", mean(overlap_meme_above_summary), "+/-", sd(overlap_meme_above_summary)))
# [1] "Average number of times more significant CTCF sites overlap with experimental CTCF sites 120.151631212219 +/- 81.188988520227"

# Among less significant CTCF sites, XX% did not overlap any cell type-specific CTCF sites.
# In contrast, only YY% more significant sites overlap
print(paste("The proportion of less significant CTCF sites overlapping at least one experimental CTCF site", 1 - (sum(overlap_meme_below_summary == 0) / length(overlap_meme_below_summary)) ))
# [1] "The proportion of less significant CTCF sites overlapping at least one experimental CTCF site 0.218351650508075"
print(paste("The proportion of more significant CTCF sites overlapping at least one experimental CTCF site", 1 - (sum(overlap_meme_above_summary == 0) / length(overlap_meme_above_summary)) ))
# [1] "The proportion of more significant CTCF sites overlapping at least one experimental CTCF site 0.888007013981819"

# P-value of the differences
res <- wilcox.test(overlap_meme_above_summary, overlap_meme_below_summary)
res$p.value
```
